Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-01-31 paper/main.tex
Updates

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remains limited, as researchers often prefer to construct DAGs manually
based on their expert knowledge. This preference arises due to several
practical issues with these automated algorithms, such as making obvious errors
and outputting Markov Equivalence Classes (MECs). To address these challenges,
we propose an iterative method to assist researchers in manually constructing
or modifying DAGs by combining implied Conditional Independence (CI) testing
with expert knowledge. We introduce a novel measure of partial association for
mixed data to rank violations of implied CIs, enabling users to prioritize
model modifications while leveraging their expertise to decide the orientation
of edges when making these modifications. Empirical results show that this
approach achieves performance comparable to automated algorithms, assuming the
user can correctly orient edges in one out of three cases. Additionally, we
demonstrate that, in the absence of an expert, a Large Language Model can
potentially determine edge orientations.

@@

In the DAG literature, the primary focus has been on developing automated
algorithms to learn causal structures from datasets. These efforts have led to
numerous causal discovery algorithms, including constraint-based methods like
the PC algorithm \citep{Spirtes2001} and Fast Causal Inference (FCI)
\citep{Spirtes2000}), score-based methods such as Hill-Climb Search and Greedy
Equivalence Search (GES) \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. In contrast, the SEM literature has focused on developing
tools to assist researchers in manually constructing or modifying models,
enabling experts to incorporate their domain knowledge in the model building
process. Researchers typically begin with an initial model based on their
expertise and then use these tools to guide modifications that improve the
model's fit to data. This process is commonly known as \emph{Specification
Search} \citep{Long1983} and uses method such as modification indices, and
Wald-based tests \citep{Marcoulides2018}. While there has been some effort in
DAG literature as well towards allowing for expert knowledge but these methods
focus towards providing the expert knowledge to the automated algorithms
\citep{Bang2023, Meek1995}. \todo[inline]{Whether to keep this sentence?}

@@

\begin{enumerate}
	\item \textbf{Bounded: } $ \rho $ is bounded between $ 0 $ and $ 1 $,
		with $ 0 $ representing no association while $ 1 $ represents
		perfect linear association. This helps in interpretation and
		comparing values across multiple partial associations.
	\item \textbf{Independent of Sample Size: } The measure remains consistent
		regardless of the sample size ensuring reliability across 
		studies with different sample sizes.
	\item \textbf{Invariant to the Dummy Encoding Scheme: } Since canonical
		correlations search for linear combinations over the columns,
		the measure is invariant to the specific dummy encoding
		scheme used for categorical variables.
	\item \textbf{Generalizes Partial Correlations to Mixed Data: }
		Canonical correlations is equivalent to Pearson's correlation coefficient
		when both $ X $ and $ Y $ are continuous. Hence, $ \rho $ generalizes
		partial correlations to mixed data types.
	\item \textbf{Generalizes polychoric and polyserial correlation: }
		Under the assumption that the ordinal variable is generated by
		discretizing an underlying gaussian continuous variable, this
		effect size is equivalent to polychoric and polyserial
		correlation. As both of them recover the Pearson correlation
		coefficient.
	\item \textbf{Related to Cramer\'s V: } 
\end{enumerate}

>>>>

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on their domain knowledge. This preference arises due to several
practical issues with automated algorithms, such as they make obvious errors
and out Markov Equivalence Classes (MECs). To address these challenges, we
propose an iterative method to assist researchers in manually constructing or
improving an existing DAG by combining implied Conditional Independence (CI)
testing with expert knowledge. Our proposed method utilizes a novel measure of
partial association for mixed data to rank violations of implied CIs in a given
DAG, enabling users to prioritize modifications while utilizing their expertise
when making these modifications. Empirical results show that this manual model
construction approach achieves performance comparable to automated algorithms
even if the user is able to decide the orientation of an edge correctly in one
out of three cases. Additionally, we demonstrate that, in the absence of an
expert, a Large Language Model (LLM) can potentially determine edge orientations.

@@

In the DAG literature, the primary focus has been on developing automated
algorithms to learn causal structures from datasets. These efforts have led to
numerous causal discovery algorithms, including constraint-based methods like
the PC algorithm \citep{Spirtes2001} and Fast Causal Inference (FCI)
\citep{Spirtes2000}), score-based methods such as Hill-Climb Search and Greedy
Equivalence Search (GES) \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. In contrast, the SEM literature has focused on developing
tools to assist researchers in manually constructing or modifying models,
enabling experts to incorporate their domain knowledge in the model building
process. Researchers typically begin with an initial model based on their
expertise and then use these tools to guide modifications that improve the
model's fit to data. This process is commonly known as \emph{Specification
Search} \citep{Long1983} and uses method such as modification indices, and
Wald-based tests \citep{Marcoulides2018}.

While there has been some effort in
DAG literature as well towards allowing for expert knowledge but these methods
focus towards providing the expert knowledge to the automated algorithms
\citep{Bang2023, Meek1995}.

@@

\begin{enumerate}
	\item \textbf{Bounded: } $ \rho $ is bounded between $ 0 $ and $ 1 $,
		with $ 0 $ representing no association while $ 1 $ represents
		perfect linear association. This helps in interpretation and
		comparing values across multiple partial associations.
	\item \textbf{Independent of Sample Size: } The measure remains consistent
		regardless of the sample size ensuring reliability across 
		studies with different sample sizes.
	\item \textbf{Invariant to the Dummy Encoding Scheme: } Since canonical
		correlations search for linear combinations over the columns,
		the measure is invariant to the specific dummy encoding
		scheme used for categorical variables.
	\item \textbf{Equivalent to Partial Correlations for Continuous variables}
		Canonical correlations is equivalent to Pearson's correlation coefficient
		when both $ X $ and $ Y $ are continuous. Hence, $ \rho $ generalizes
		partial correlations to mixed data types.
	\item \textbf{Equivalent to polychoric and polyserial correlation for ordinal and continuous variables: }
		Under the assumption that the ordinal variable is generated by
		discretizing an underlying gaussian continuous variable, this
		effect size is equivalent to polychoric and polyserial
		correlation. As both of them recover the Pearson correlation
		coefficient.
	\item \textbf{Related to Cramer\'s V: } 
\end{enumerate}
